Quantum H theorem, variational entropy, and relaxation processes

### 2020-10-21 notasparaarticulo.tex
the tex. file was checked on grammarly, some comments were included in chapter 7 as a footnote, the abstract is still unchanged

\textcolor{red}{In the literature, the theoretical basis to describe equilibrium systems in the classical scheme is well-established. From thermodynamics to statistical mechanics, the procedures to describe mathematically equilibrium systems for both schemes (phenomenological for thermodynamics and microscopical for statistical mechanics) are well-known and works in almost every system in nature \cite{huang, reif, patrick}. In the case of the out of equilibrium systems, there are many works that attempt to describe out of equilibrium systems (however those works do not work in some systems, for example, the internal behavior of a glass due to the time relaxation is very big \cite{cristal1} and the entropy measurement \cite{cristal2, cristal3} as examples). Examples of some theoretical schemes to describe out of equilibrium systems can be the Onsager formulation (called linear thermodynamics), some few other restricted developments that attempt to complete the Onsager formulation \cite{kei, onsager} and the Kinetic Theory of dilute gases developed by Boltzmann.\\
In the case of the Kinetic Theory of dilute gases, the behavior of a dilute classical gas was described by Boltzmann's by means the Boltzmann transport equation. Furthermore, the evolution to an equilibrium state for any system was proved by means the Boltzmann's $H$ theorem (or classical $H$ theorem) and the spatial-homogeneous distribution function hypothesis.
However, there are some some aspects in the Boltzmann's $H$ theorem and the Boltzmann's $H$ functional that deserve some review in order to accomplish better consistency, for example the modification of the $H$ theorem in stochastic trajectories, violations in the second law of thermodynamics, and the relation between Shannon's measure of information and the Boltzmann's entropy  [6-10].\\ %\cite{paradox1, paradox2, hamiltonian, H-theorem-violation, H-theorem-and-entropy}.

@@

In order to obtain the well-known correspondence principle between classical mechanics and quantum mechanics, a quantum version of the $H$ theorem and the Boltzmann transport equation are necessary such that in the appropriate limit, the  Boltzmann's $H$ theorem and the Boltzmann transport equation should be recovered.\\
One of the first developments in this field was developed by Tolman \cite{tolman}. The quantum version of the $H$ theorem was established by Tolman using the probability transition relation, the random phases hypothesis and a homogeneous distribution function. The equation of motion for the occupation number was obtained by applying the time perturbation theory. However, in order to obtain the flux relations in analogy to the Boltzmann's method, a quantum version of the transport equation that includes a non-homogeneous distribution function is required.\\
In other hand, some works in the quantum operator formalism attempt to describe quantum transport phenomena using the Hamiltonian of the system and the master equation (the analogous to the Boltzmann transport equation) [12-15]. %\cite{entropic-framework, thermal-harmonic, quantum-transport, quantum-transport-reservoirs}.
Besides some papers construct the $H$ functional and prove the $H$ theorem \cite{htheorem2, quantum1, quantum2}. However, the homogeneous distribution function hypothesis and the correspondence principle are not clear or discussed. Moreover the validity of the quantum $H$ theorem and the second low of thermodynamics, and the interpretation of the quantum entropy is still in discussion [19-26].\\ %\cite{quantum-entropy, quantum-colapse, quantum3, quantum4, binary, contradictions,ergodic-theorem, hydrodynamic-simulation}.\\

@@

In order to contribute to a better consistent $H$ theorem and a new formalism to describe out of equilibrium systems, we define the theoretical basis to describe out of equilibrium systems in order to be completed in the future.\\
In this work we define a new functional (variational entropy) which depends on the distribution function of an ideal quantum gas (Bosons or Fermions particles). This gas is contained inside a volume $V$ divided into cells. Before developing the non-equilibrium description, we recover the Fermi-Dirac and Bose-Einstein distributions from the variational entropy when the system is in the equilibrium state. Furthermore we prove that the derivative of the variational entropy with respect to time is less or equal to zero for relaxation processes (the action of any perturbed system that returns to the equilibrium state) in any system. This prove corresponds with the statement of the $H$ theorem. In addition we obtain the Boltzmann transport equation from the Boltzmann's $H$ functional and, in analogy with this, we find a quantum version of the Boltzmann transport equation from the variational entropy. Finally, in order to be consistence with the correspondence principle, we prove that the Boltzmann's $H$ functional is proportional to the variational entropy in the limit of the high level energies.}

@@

\section{The equilibrium of dilute gases: a review}
\subsection{Classical approach}
In the classical version, the Boltzmann transport equation is used to shown the $H$-theorem. That equation is
\begin{equation}
    \left( \frac{\partial }{\partial t}+v_1 \cdot \nabla_r +\frac{F}{m} \cdot \nabla_{v_1} \right)f_1=\int \mathrm{d}\Omega \int \mathrm{d}^{3}v_2\sigma(\Omega)|v_1-v_2|(f_2'f_1'-f_2f_1),
\end{equation}
where $\Omega$ is the solid angle, $\sigma$ is the scattering section, $F$ the external force applied to the system, $f_1, f_2$ are the distribution function of the particles 1 and 2 respectively before the collision, and $f'_1, f'_2$ are the distribution function of the particles 1 and 2 respectively after the collision.\\
The $H$ theorem can be proved using the $H$ functional defined by Boltzmann 
\begin{equation}
    H=\int f(v,t) \ln f(v,t) \mathrm{d}^{3}v,
\end{equation} \label{hfunctional}
where $f(v,t)$ is a homogeneous distribution function. The $H$ theorem establishes that if $f(v,t)$ satisfies the Boltzmann transport equation, then $\frac{dH}{dt}\leq 0$. The system arrives to the equilibrium state when $\frac{dH}{dt}=0$ and due to this, the distribution function to the system in equilibrium corresponds to the Maxwell-Boltzmann distribution
\subsection{Quantum approach}
The $H$ functional in quantum mechanics presented by Tolman \cite{tolman} is defined as 
\begin{equation}
    H=\sum_i n_i \ln n_i -(n_i\pm g_i)\ln (g_i \pm n_i)\pm g_i\ln g_i, \label{quantumh}
\end{equation}
where $n_i$ is the occupation number and $g_i$ is a energy cell. Also, an equation for the rate of change in the number of particles in group $\kappa$ is
\begin{eqnarray}
    \frac{d n_{\kappa}}{dt}&=&-\sum_{\lambda,(\mu \nu)}A_{\kappa\lambda,\mu\nu} n_{\kappa}n_{\lambda}(g_{\mu}\pm n_{\mu})(g_{\nu}\pm n_{\nu})\nonumber \\
    &&+\sum_{\lambda,(\mu \nu)}A_{\mu\nu,\kappa\lambda} n_{\mu}n_{\nu}(g_{\kappa}\pm n_{\kappa})(g_{\lambda}\pm n_{\lambda}),\label{changen}
\end{eqnarray}
where $A_{\kappa\lambda,\mu\nu}$ is a tensor expressed as
\begin{equation}
  A_{\kappa\lambda,\mu\nu}=\frac{4\pi^{2}}{h}\frac{|I_1\pm I_2|^2}{\Delta \epsilon},
\end{equation}
where $\epsilon$ is the energy and $|I_1-I_2|^2=|V_{mn,kl}|^2$ where $V_{mn,kl}$ is the perturbation matrix of binary particles. Those expression are obtained by the perturbation theory.\\
The quantum $H$-theorem is proved using (\ref{changen}) into (\ref{quantumh}), taking account the spatial-homogeneous distribution function hypothesis, as in the classical version.\\
In addition to this, Tolman showed that (\ref{quantumh}) can be reduced in the limit of high energy levels to
\begin{equation}
    H = \sum_{\kappa} (n_{\kappa} \ln n_{\kappa} - n_{\kappa} \ln g_{\kappa}). \label{reduce-h}
\end{equation}
This expression can be obtained also from the Boltzmann $H$ functional by means defining 
\begin{equation}
    f=\frac{n_{\kappa}}{ g_{\kappa}}.
\end{equation}
\section{Ensemble approach}
\subsection{Quantum scheme}

@@

This functional, in equilibrium should be a density of entropy, which we will call \textit{variational entropy}.
This functional describes a system with global energy E, global number of free
quantum particles (which may be fermions for upper sing or bosons for lower sing) N, global volume V, and temperature T. The
global system in turn consists of a set of $K$ small cells with a constant volume $\delta V_M$, whose local homogeneous distribution functions is $f_{Mn}(\epsilon_{n},t)$. The distribution function will depend on the position of the cells, the energy per cell $\epsilon_{n}$, and time $t$. Besides the energy spectra will be divided into $\delta \epsilon_n$ volumes. We will use the uppercase latin index to denote cell numbers, and unless stated otherwise, $M = 1, 2 . . . , K$. Without loss of generality, we choose these cells to have
local identical volumes (\textit{i.e.} $\delta V_M = V/K, \forall M$). Each cell has its own local chemical potential, $\mu_M$, local

@@

therefore the energy spectrum, denoted by $\epsilon_n$, is the same for all cells, \textit{i.e.} $\epsilon_{Mn} = \epsilon_n, \forall M$. In addition, the system is subjected to the microcanonical restrictions

@@

We proved, when the system is in equilibrium, the Lagrange multipliers don't depend on the cell number. In other words, Lagrange multiplier are homogeneous in the system.\\
We can calculate the entropy from an ideal system, in other words, if one substitute (\ref{distributionequilibrium}) to (\ref{entropy}) one obtains

@@

  This result lets us calculate the entropy of a quantum ideal gases. This entropy could be used as a thermodynamic variable to describe quantum ideal gases.

@@

\begin{equation}
   H'(t)=\sum_M\int_{}^{} f_M(v,t) \ln f_M(v,t)\mathrm{d}^3v  \label{CH2}.
\end{equation}
The functional (\ref{CH2}) contains a no-homogeneous distribution function but keeps the same form of the Boltzman's $H$ functional. Excluding the homogeneous distribution function hypothesis, we use the same procedure that it was shown in the previous section to prove the $H$-theorem.\\ 
If one considers the microcanonical ensemble conditions 
\begin{equation}
    \sum_M\int_{}^{}f_M(v,t)\mathrm{d}^3v =N, \ \ \ \sum_M\int_{}^{}f_M(v,t)\epsilon(v)\mathrm{d}^3v =E \label{micro},
\end{equation}{}
one obtains the distribution function that minimizes $H$
\begin{eqnarray}
\frac{\delta H'}{\delta f_J(v')}&=&\sum_M\int_{}^{}\frac{\delta}{\delta f_J(v')}\left[f_M(v)\ln f_M(v)  \right]\mathrm{d}^3v -\sum_M\alpha_M\int_{}^{}\frac{\delta f_M(v)}{\delta f_J(v')}\mathrm{d}^3v \nonumber \\
&&-\sum_M\beta_M \int_{}^{}\epsilon(v)\frac{\delta f_M(v)}{\delta f_J(v')}\mathrm{d}^3v \nonumber \\
&=&\ln f_J(v')+1-\alpha_J-\beta_J \epsilon(v')=0.
\end{eqnarray}{}
isolating the distribution function 
\begin{eqnarray}
\ln f(r',v')&=&\alpha(r')+\beta(r') \epsilon(v')-1 \ \ \  \Rightarrow \ \ \ f_J(v')=e^{\alpha_J +\beta_J \epsilon(v')-1} \nonumber \\
&=&Ce^{\alpha_J+\beta_J \epsilon(v') } \label{relacion1},
\end{eqnarray}{}
where $C$ is a constant. It is trivial that using the variational procedures, one obtains the form of the Maxwell-Boltzmann distribution. \\
As $H'$ in equilibrium \textcolor{red}{is proportional to} entropy, it is trivial to think that Lagrange multipliers do not depend on the position because of the system is in equilibrium and consequently, the distribution function is homogeneous, that is
\begin{equation}
    f(v)=Ce^{\alpha+\beta \epsilon(v)}.
\end{equation}{} 
The constant $C$ can be omitted defining the following $H$ functional
\begin{equation}
   H''(t)=\sum_M\int_{}^{} \left[f_M(v,t) \ln f_M(v,t)-f_M(v,t)\right]\mathrm{d}^3v  \label{CH3},
\end{equation}{}
  but $\sum_M \int f_M(v,t)\mathrm{d}^3v =N$ where $N$ is the total particle number. As $N$ is a constant and any $H$ functional has sense when we calculate its time derivative, this constant can be omitted. \\
  \textcolor{green}{In other hand, if the distribution function is spatial-homogeneous, $f$ does not depend on the cell number $M$, then we rewrite (\ref{CH2}) as
  \begin{equation}
      H'(t)=\int \sum_M [f(v,t)\ln f(v,t)] \mathrm{d}^{3}v = P\int  f(v,t) \ln f(v,t) \mathrm{d}^{3}v= P H_{boltz}(t), 
  \end{equation}
  where
  \begin{equation}
      H_{boltz}(t)=\int  f(v,t) \ln f(v,t) \mathrm{d}^{3}v.
  \end{equation}
  Besides, if we analyse the following expression
  \begin{eqnarray}
      \int Pf(v,t) \ln [Pf(v,t)] \mathrm{d}^{3}v = \int [(P\ln P)f(v,t) + Pf(v,t) \ln f(v,t)]\mathrm{d}^{3}v, \label{sum-h}
  \end{eqnarray}
  we can observe if the number of particles in the $\mu$-space $f(v,t)$ is to big compare with the number of cells $P$, then the the first term in (\ref{sum-h}) is negible and and consequently we conclude that
  \begin{equation}
      \int f'(v,t) \ln f'(v,t) \mathrm{d}^{3}v = P\int f(v,t) \ln f(v,t) \mathrm{d}^{3}v; \ \ \ f'(v,t)= Pf(v,t). \label{aditive-property} 
  \end{equation}
  We obtained the additive property when we supposed the distribution function is spatial-homogeneous.}

@@

The \textit{variational entropy} has the following form
\begin{equation}
    \Ss(t)=\sum_{M} \sum_{n} \left[ f_{Mn} \ln f_{Mn} \pm (1 \mp f_{Mn}) \ln (1 \mp f_{Mn}) \right] \de V_M \delta \epsilon_n \label{entropy2},
\end{equation}
where, for simplicity, we have defined $f(\epsilon_{n},t)\equiv f_{Mn}$.
This distribution function \textcolor{red}{must satisfy the local equilibrium hypothesis in each cell ($f_{nM}$ must be spatial-homogeneous for each $M$), the non-homogeneous distribution function assumption in the total volume ($f_{nM}$ must be different among all cells)}, and the following microcanonical restrictions
\begin{eqnarray}
        \sum_{n}f_{Mn} \delta \epsilon_n=\bar{\mathcal{N}}+\Delta_M(t); \ \ \ \ \sum_{n}\epsilon_{n}f_{Mn} \delta \epsilon_n=\bar{\mathcal{E}}+ \delta_M(t), \label{restrictionoutside}
  \end{eqnarray}
  where $\bar {\mathcal{N}}$ and $\bar{\mathcal{E}}$ are the local particle number and the local energy in equilibrium
  \begin{equation}
      \bar{\mathcal{N}}= \sum_n \bar{f}_n \delta \epsilon_n; \ \ \ \ \bar{\mathcal{E}}= \sum_n \epsilon_n\bar{f}_n \delta \epsilon_n,
  \end{equation}
  and $\Delta_M,\delta_M$ could be seen as a deviation from $\bar{\mathcal{N}}$ and $\bar{\mathcal{E}}$ respectively, with $\Delta_M(t)\ll \bar{\mathcal{N}}$ and $\delta_M(t) \ll \bar{\mathcal{E}}$. Those condition described a system out of equilibrium but no so far from it, that is a perturbed system.  \\
If we perform the derivative of the variational entropy with respect to time, we obtain
\begin{equation}
   \frac{d \Ss (t)}{dt}= \sum_n \sum_M \dot{f}_{nM}(t)\ln \left[ \frac{f_{nM}(t)}{1\mp f_{nM}(t)} \right] \de V_M \delta \epsilon_n.\label{deltaH}
\end{equation}{}
Using the first-order approximation
\begin{eqnarray}
   f_{Mn}=\bar{f}_{n}(1+g_{Mn}): \ \ \ \bar{f}_{n}\gg \bar{f}_{n}|g_{Mn}|, \label{firstorder}
\end{eqnarray}{}
we get

@@

The previous expression establishes that the value of $g_{nM}$ is determined by $\bar f_n$. Furthermore, whether $\bar f_{n}=1$, this is, all lower energy levels are occupied, then the system will not have inhomogenities because of the exclusion principle and as consequence of this,  $g_{nM}=0$. This result corresponds to the system in the zero temperature condition. We can observe that in this condition, (\ref{fermionrestriction}) is violated. Due to this, the zero temperature condition is excluded.\\

@@

On the one hand, $\dot{g}^{+}_{nJ}<0$ describes the lost of particles and/or energy and so , $g^{+}_{nJ}>0$. On the other hand, $\dot{g}^{-}_{nJ}>0$ describes the gain of particles and/or energy and therefore $g^{-}_{nJ}<0$. \\

@@

When the system is in equilibrium, $g_{nM}=0$ and in consequence of that, $\dot g_{nM}=0$ and finally from (\ref{cambioH1}) $\frac{d\Ss}{dt}=0$. Then, we can say that any quantum ideal gas perturbed, evolves always to the equilibrium state, such as the classical $H$ theorem. \\
Now we will apply this method to the modified Boltzmann's $H$ functional (\ref{CH2}) to prove the $H$ theorem without using only the first order approximation (\ref{firstorder}), and the behavior of the deviation $g_{nM}$ and its time derivative (\ref{separacion}).

@@

\section{Conclusions}
\textcolor{red}{We conclude the following statements:\\
\begin{enumerate}
    \item Any system (classical or quantum) out of equilibrium (including inhomogeneities) evolves to the equilibrium state without using the transport equation as long as the local equilibrium hypothesis is hold.
    \item We can obtain the transport equation from the $H$ functional.
    \item The variational entropy is proportional to the Boltzmann's $H$ functional in the limit of the high energy levels.
    \item We find an expression for the entropy to quantum gases.
    \end{enumerate}
  }
%  \begin{equation}
%      \mathcal{H}
%  \end{equation}

>>>>

\textcolor{red}{In the literature, the theoretical basis that allows us to describe equilibrium systems in the classical scheme is well-established. From thermodynamics to statistical mechanics, the procedures to describe mathematically equilibrium systems for both schemes (phenomenological for thermodynamics and microscopical for statistical mechanics) are well-known and works in almost every system in nature. In the case of the out of equilibrium systems, many physics attempt to describe out of equilibrium systems (however those works do not work in some systems, for example, the internal behavior of glasses due to the time relaxation is  \cite{huang, reif, patrick}. In the case of the out of equilibrium systems, many physics attempt to describe out of equilibrium systems creating new theoretical basis (however those works do not work in some systems, for example, the internal behavior of glasses due to the time relaxation is very big \cite{cristal1} and the entropy measurement \cite{cristal2, cristal3} as examples). Examples of some theoretical basis can be the Onsager formulation (called linear thermodynamics), some few other restricted developments that attempt to complete the Onsager formulation \cite{kei, onsager} and the Kinetic Theory of dilute gases developed by Boltzmann.\\
In the case of the Kinetic Theory of dilute gases, the behavior of a dilute classical gas was described by Boltzmann's through the Boltzmann transport equation. Furthermore, the evolution to an equilibrium state for any system was proved by means the Boltzmann's $H$ theorem (or classical $H$ theorem) and the spatial-homogeneous distribution function hypothesis.
However, there are some aspects in the Boltzmann's $H$ theorem and the Boltzmann's $H$ functional that deserve some review to accomplish better consistency, for example, the modification of the $H$ theorem in stochastic trajectories, violations in the second law of thermodynamics, and the relation between Shannon's measure of information and the Boltzmann's entropy  [6-10].\\ %\cite{paradox1, paradox2, hamiltonian, H-theorem-violation, H-theorem-and-entropy}.

@@

To obtain the well-known correspondence principle between classical mechanics and quantum mechanics, a quantum version of the $H$ theorem and the Boltzmann transport equation is necessary such that in the appropriate limit, the  Boltzmann's $H$ theorem and the Boltzmann transport equation should be recovered.\\
One of the first developments in this field was developed by Tolman \cite{tolman}. He proposes one of the first quantum versions of the $H$ theorem using the probability transition relation, the random phases hypothesis, and a homogeneous distribution function. The equation of motion for the occupation number (necessary to prove the quantum $H$ theorem in the same way that Boltzmann's procedure) was obtained by applying the time perturbation theory. However, to obtain the flux relations in analogy to the Boltzmann's method, a quantum version of the transport equation that includes a non-homogeneous distribution function is required.\\
On the other hand, some works in the quantum operator formalism attempt to describe quantum transport phenomena using the Hamiltonian of the system and the master equation (the analogous to the Boltzmann transport equation) [12-15]. %\cite{entropic-framework, thermal-harmonic, quantum-transport, quantum-transport-reservoirs}.
Besides, some authors construct the $H$ functional and prove the $H$ theorem \cite{htheorem2, quantum1, quantum2}. However, the homogeneous distribution function hypothesis and the correspondence principle are uncleared or discussed. Moreover, the validity of the quantum $H$ theorem and the second law of thermodynamics, and the interpretation of the quantum entropy is still in discussion [19-26].\\ %\cite{quantum-entropy, quantum-colapse, quantum3, quantum4, binary, contradictions,ergodic-theorem, hydrodynamic-simulation}.\\

@@

In order to contribute to a better consistent $H$ theorem and a new formalism to describe out of equilibrium systems, we define the theoretical basis to describe out of equilibrium systems.\\
In this work, we define a new functional (variational entropy) which depends on the distribution function of an ideal quantum gas (Bosons or Fermions). This gas is contained inside a volume $V$ divided into cells. Before developing the non-equilibrium description, we recover the Fermi-Dirac and Bose-Einstein distributions from the variational entropy when the system is in the equilibrium state. Furthermore, we prove that the derivative of the variational entropy with respect to time is less or equal to zero for relaxation processes (the action of any perturbed system that returns to the equilibrium state) in any system. This prove corresponds with the statement of the $H$ theorem. Besides, we obtain the Boltzmann transport equation from the Boltzmann's $H$ functional and, in analogy with this, we find a quantum version of the Boltzmann transport equation from the variational entropy. Finally, to be consistent with the correspondence principle, we prove that the Boltzmann's $H$ functional is proportional to the variational entropy in the limit of the high-level energies.}

@@

\section{The equilibrium of dilute gases: a review}
\subsection{Classical approach}
In the classical version, the Boltzmann transport equation is required to show the $H$-theorem. That equation is
\begin{equation}
    \left( \frac{\partial }{\partial t}+v_1 \cdot \nabla_r +\frac{F}{m} \cdot \nabla_{v_1} \right)f_1=\int \mathrm{d}\Omega \int \mathrm{d}^{3}v_2\sigma(\Omega)|v_1-v_2|(f_2'f_1'-f_2f_1),
\end{equation}
where $\Omega$ is the solid angle, $\sigma$ is the scattering section, $F$ the external force applied to the system, $f_1, f_2$ are the distribution function of the particles 1 and 2 respectively before the collision, and $f'_1, f'_2$ are the distribution function of the particles 1 and 2 respectively after the collision.\\
The $H$ theorem can be proved using the $H$ functional defined by Boltzmann 
\begin{equation}
    H=\int f(v,t) \ln f(v,t) \mathrm{d}^{3}v,\label{hfunctional}
\end{equation} 
where $f(v,t)$ is a homogeneous distribution function. The $H$ theorem establishes that if $f(v,t)$ satisfies the Boltzmann transport equation, then $\frac{dH}{dt}\leq 0$. The system arrives to the equilibrium state when $\frac{dH}{dt}=0$, and due to this the distribution function to the system in equilibrium corresponds to the Maxwell-Boltzmann distribution
\subsection{Quantum approach}
The $H$ functional in quantum mechanics presented by Tolman \cite{tolman} is defined as 
\begin{equation}
    H=\sum_i n_i \ln n_i -(n_i\pm g_i)\ln (g_i \pm n_i)\pm g_i\ln g_i, \label{quantumh}
\end{equation}
where $n_i$ is the occupation number and $g_i$ is an energy cell. Also, an equation for the rate of change in the number of particles in group $\kappa$ is
\begin{eqnarray}
    \frac{d n_{\kappa}}{dt}&=&-\sum_{\lambda,(\mu \nu)}A_{\kappa\lambda,\mu\nu} n_{\kappa}n_{\lambda}(g_{\mu}\pm n_{\mu})(g_{\nu}\pm n_{\nu})\nonumber \\
    &&+\sum_{\lambda,(\mu \nu)}A_{\mu\nu,\kappa\lambda} n_{\mu}n_{\nu}(g_{\kappa}\pm n_{\kappa})(g_{\lambda}\pm n_{\lambda}),\label{changen}
\end{eqnarray}
where $A_{\kappa\lambda,\mu\nu}$ is a tensor expressed as
\begin{equation}
  A_{\kappa\lambda,\mu\nu}=\frac{4\pi^{2}}{h}\frac{|I_1\pm I_2|^2}{\Delta \epsilon},
\end{equation}
where $\epsilon$ is the energy and $|I_1-I_2|^2=|V_{mn,kl}|^2$ where $V_{mn,kl}$ is the perturbation matrix of binary particles. Those expressions are obtained by the perturbation theory.\\
The quantum $H$-theorem is proved using (\ref{changen}) into (\ref{quantumh}), taking account of the spatial-homogeneous distribution function hypothesis, as in the classical version.\\
In addition to this, Tolman showed that (\ref{quantumh}) can be reduced in the limit of high-energy levels to
\begin{equation}
    H = \sum_{\kappa} (n_{\kappa} \ln n_{\kappa} - n_{\kappa} \ln g_{\kappa}). \label{reduce-h}
\end{equation}
This expression also can be obtained also from the Boltzmann $H$ functional through defining 
\begin{equation}
    f=\frac{n_{\kappa}}{ g_{\kappa}}.
\end{equation}
\section{Ensemble approach}
\subsection{Quantum scheme}

@@

This functional will be named \textit{variational entropy}.
This functional describes a system with a global energy E, a global number of free
quantum particles (which may be fermions for upper sing or bosons for lower sing) N, global volume V, and a temperature T. 
Besides, the variational entropy should correspond to a density of entropy when the system arrives at the equilibrium state. The
global system, in turn, consists of a set of $K$ small cells with a constant volume $\delta V_M$, whose local homogeneous distribution functions is $f_{Mn}(\epsilon_{n},t)$. The distribution function will depend on the position of the cells, the energy per cell $\epsilon_{n}$, and time $t$. Also, the energy spectra will be divided into $\delta \epsilon_n$ volumes. We will use the uppercase Latin index to denote cell numbers, and unless stated otherwise, $M = 1, 2 . . . , K$. Without loss of generality, we choose these cells to have
local identical volumes (\textit{i.e.} $\delta V_M = V/K, \forall M$). Each cell has its local chemical potential, $\mu_M$, local

@@

therefore the energy spectrum, denoted by $\epsilon_n$, is the same for all cells, \textit{i.e.} $\epsilon_{Mn} = \epsilon_n, \forall M$. Besides, the system is subjected to the microcanonical restrictions

@@

We proved, when the system is in equilibrium, the Lagrange multipliers don't depend on the cell number. In other words, Lagrange multipliers are homogeneous in the system.\\
We can calculate the entropy from an ideal system, in other words, if one substitute (\ref{distributionequilibrium}) to (\ref{entropy}) we obtain

@@

  This result allows us to calculate the entropy of a quantum ideal gases. This entropy could be used as a thermodynamic variable to describe those systems.

@@

\begin{equation}
   H'(t)=\sum_M\int_{}^{} f_M(v,t) \ln f_M(v,t)\mathrm{d}^3v  \label{CH2}.
\end{equation}
The functional (\ref{CH2}) contains a no-homogeneous distribution function but keeps the same form of the Boltzman's $H$ functional. Excluding the homogeneous distribution function hypothesis, we use the same procedure that it was shown in the previous section to prove the $H$-theorem.\\ 
If one considers the microcanonical ensemble conditions 
\begin{equation}
    \sum_M\int_{}^{}f_M(v,t)\mathrm{d}^3v =N, \ \ \ \sum_M\int_{}^{}f_M(v,t)\epsilon(v)\mathrm{d}^3v =E \label{micro},
\end{equation}{}
one obtains the distribution function that minimizes $H$
\begin{eqnarray}
\frac{\delta H'}{\delta f_J(v')}&=&\sum_M\int_{}^{}\frac{\delta}{\delta f_J(v')}\left[f_M(v)\ln f_M(v)  \right]\mathrm{d}^3v -\sum_M\alpha_M\int_{}^{}\frac{\delta f_M(v)}{\delta f_J(v')}\mathrm{d}^3v \nonumber \\
&&-\sum_M\beta_M \int_{}^{}\epsilon(v)\frac{\delta f_M(v)}{\delta f_J(v')}\mathrm{d}^3v \nonumber \\
&=&\ln f_J(v')+1-\alpha_J-\beta_J \epsilon(v')=0.
\end{eqnarray}{}
isolating the distribution function 
\begin{eqnarray}
\ln f(r',v')&=&\alpha(r')+\beta(r') \epsilon(v')-1 \ \ \  \Rightarrow \ \ \ f_J(v')=e^{\alpha_J +\beta_J \epsilon(v')-1} \nonumber \\
&=&Ce^{\alpha_J+\beta_J \epsilon(v') } \label{relacion1},
\end{eqnarray}{}
where $C$ is a constant. It is trivial that using the variational procedures we obtain the form of the Maxwell-Boltzmann distribution. \\
As $H'$ in equilibrium \textcolor{red}{is proportional to} entropy, it is trivial to think that Lagrange multipliers do not depend on the position because of the system is in equilibrium and consequently, the distribution function is homogeneous, that is
\begin{equation}
    f(v)=Ce^{\alpha+\beta \epsilon(v)}.
\end{equation}{} 
The constant $C$ can be omitted defining the following $H$ functional
\begin{equation}
   H''(t)=\sum_M\int_{}^{} \left[f_M(v,t) \ln f_M(v,t)-f_M(v,t)\right]\mathrm{d}^3v  \label{CH3},
\end{equation}{}
  but $\sum_M \int f_M(v,t)\mathrm{d}^3v =N$ where $N$ is the total particle number. As $N$ is a constant and any $H$ functional has sense when we calculate its time derivative, this constant can be omitted. \\
  \textcolor{gray}{In other hand, if the distribution function is spatial-homogeneous, $f$ does not depend on the cell number $M$, then we rewrite (\ref{CH2}) as
  \begin{equation}
      H'(t)=\int \sum_M [f(v,t)\ln f(v,t)] \mathrm{d}^{3}v = P\int  f(v,t) \ln f(v,t) \mathrm{d}^{3}v= P H_{boltz}(t), 
  \end{equation}
  where
  \begin{equation}
      H_{boltz}(t)=\int  f(v,t) \ln f(v,t) \mathrm{d}^{3}v.
  \end{equation}
  Besides, if we analyze the following expression
  \begin{eqnarray}
      \int Pf(v,t) \ln [Pf(v,t)] \mathrm{d}^{3}v = \int [(P\ln P)f(v,t) + Pf(v,t) \ln f(v,t)]\mathrm{d}^{3}v, \label{sum-h}
  \end{eqnarray}
  we can observe if the number of particles in the $\mu$-space $f(v,t)$ is to big compare with the number of cells $P$, then the first term in (\ref{sum-h}) is negligible, and consequently, we conclude that
  \begin{equation}
      \int f'(v,t) \ln f'(v,t) \mathrm{d}^{3}v = P\int f(v,t) \ln f(v,t) \mathrm{d}^{3}v; \ \ \ f'(v,t)= Pf(v,t). \label{aditive-property} 
  \end{equation}
  We obtained the additive property when we supposed the distribution function is spatial-homogeneous.}

@@

The \textit{variational entropy} has the following form
\begin{equation}
    \Ss(t)=\sum_{M} \sum_{n} \left[ f_{Mn} \ln f_{Mn} \pm (1 \mp f_{Mn}) \ln (1 \mp f_{Mn}) \right] \de V_M \delta \epsilon_n \label{entropy2},
\end{equation}
where, for simplicity, we have defined $f(\epsilon_{n},t)\equiv f_{Mn}$.
This distribution function \textcolor{red}{must satisfy the local equilibrium hypothesis in each cell ($f_{nM}$ must be spatial-homogeneous for each $M$), the non-homogeneous distribution function assumption in the total volume ($f_{nM}$ must be different among all cells)}, and the following microcanonical restrictions
\begin{eqnarray}
        \sum_{n}f_{Mn} \delta \epsilon_n=\bar{\mathcal{N}}+\Delta_M(t); \ \ \ \ \sum_{n}\epsilon_{n}f_{Mn} \delta \epsilon_n=\bar{\mathcal{E}}+ \delta_M(t), \label{restrictionoutside}
  \end{eqnarray}
  where $\bar {\mathcal{N}}$ and $\bar{\mathcal{E}}$ are the local particle number and the local energy in equilibrium
  \begin{equation}
      \bar{\mathcal{N}}= \sum_n \bar{f}_n \delta \epsilon_n; \ \ \ \ \bar{\mathcal{E}}= \sum_n \epsilon_n\bar{f}_n \delta \epsilon_n,
  \end{equation}
  and $\Delta_M,\delta_M$ could be seen as a deviation from $\bar{\mathcal{N}}$ and $\bar{\mathcal{E}}$ respectively, with $\Delta_M(t)\ll \bar{\mathcal{N}}$ and $\delta_M(t) \ll \bar{\mathcal{E}}$. Those condition described a system out of equilibrium but no so far from it, that is a perturbed system.\\
  It is important to remark that the quantities $\Delta_M,\delta_M$ are sufficiently big to be different from fluctuations in the system but sufficiently small to consider the system in an out of equilibrium state but not far from the equilibrium state.  \\
If we perform the derivative of the variational entropy with respect to time, we obtain
\begin{equation}
   \frac{d \Ss (t)}{dt}= \sum_n \sum_M \dot{f}_{nM}(t)\ln \left[ \frac{f_{nM}(t)}{1\mp f_{nM}(t)} \right] \de V_M \delta \epsilon_n.\label{deltaH}
\end{equation}{}
Using the first-order approximation
\begin{eqnarray}
   f_{Mn}=\bar{f}_{n}(1+g_{Mn}): \ \ \ \bar{f}_{n}\gg \bar{f}_{n}|g_{Mn}|, \label{firstorder}
\end{eqnarray}{}
we get

@@

The previous expression establishes that the value of $g_{nM}$ is determined by $\bar f_n$. Furthermore, whether $\bar f_{n}=1$, this is, all lower energy levels are occupied, then the system will not have inhomogeneities because of the exclusion principle, and as a consequence of this,  $g_{nM}=0$. This result corresponds to the system in the zero temperature condition. We can observe that in this condition, (\ref{fermionrestriction}) is violated. Due to this, the zero temperature condition is excluded.\\

@@

On the one hand, $\dot{g}^{+}_{nJ}<0$ describes the loss of particles and/or energy and so, $g^{+}_{nJ}>0$. On the other hand, $\dot{g}^{-}_{nJ}>0$ describes the gain of particles and/or energy and therefore $g^{-}_{nJ}<0$. \\

@@

When the system is in equilibrium, $g_{nM}=0$ and in consequence of that, $\dot g_{nM}=0$ and finally from (\ref{cambioH1}) $\frac{d\Ss}{dt}=0$. Then, we can say that any ideal quantum gas perturbed always evolves to the equilibrium state, such as the classical $H$ theorem. \\
Now we will apply this method to the modified Boltzmann's $H$ functional (\ref{CH2}) to prove the $H$ theorem without using only the first-order approximation (\ref{firstorder}), and the behavior of the deviation $g_{nM}$ and its time derivative (\ref{separacion}).

@@

\section{Conclusions}
\textcolor{red}{We conclude the following statements:\\
\begin{enumerate}
    \item Any system (classical or quantum) out of equilibrium (including inhomogeneities) evolves to the equilibrium state without using the transport equation as long as the local equilibrium hypothesis holds.
    \item We can obtain the transport equation from the $H$ functional.
    \item The variational entropy is proportional to the Boltzmann's $H$ functional in the limit of the high energy levels.
    \item We find an expression for the entropy to quantum gases.
    \end{enumerate}
  }
%  \begin{equation}
%      \mathcal{H}
%  \end{equation}
